---
title: "cyCombine"
author: "Søren Helweg Dam"
date: "January 26, 2021"
output: 
  prettydoc::html_pretty:
    theme: hpstr
    highlight: vignette
vignette: >
  %\VignetteIndexEntry{cyCombine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  strip.white = T, comment = ""
)
```

**This vignette will demonstrate a general use-case of cyCombine.**

Alongside a directory of FCS files, a metadata and a panel file are assumed present.
The metadata is in this example assumed to have the following columns:

| Filename | batch | condition | Patient_id |
|---------:|:------|-----------|:----------:|

<br>

And the panel should contain the columns:

| Channel | Antigen | Type |
|---------|---------|------|

<br><br>


### Prepare data

First, the FCS files are converted into a work-able *tibble*.

#### The recommended method

```{r, eval=FALSE}
# Load packages
library(cyCombine)
library(tidyverse)

# Directory with FCS files
data_dir <- "data"

# Extract markers from panel
panel_file <- file.path(data_dir, "panel.csv") # Can also be .xlsx
metadata_file <- file.path(data_dir, "metadata.csv") # Can also be .xlsx

# Extract markers of interest
markers <- read.csv(panel_file) %>% 
  filter(Type != "none") %>% 
  pull(Antigen)

# Prepare a tibble from directory of FCS files
uncorrected <- prepare_data(
  data_dir = data_dir,
  metadata = metadata_file, 
  filename_col = "Filename",
  batch_ids = "batch",
  condition = "condition",
  down_sample = FALSE,
  markers = markers
)
# Store result
saveRDS(uncorrected, file = file.path(data_dir, "uncorrected.RDS"))
```

<br>

#### Modular method

In case you want more control, you can adjust any step of this approach.
Feel free to skip this segment, if the method above worked for you.

This example will include all possible input parameters to give an overview of what can be modified.

```{r, eval=FALSE}
# Load packages
library(cyCombine)
library(tidyverse)

# Directory with FCS files
data_dir <- "data"

# Extract markers from panel
panel <- read_csv(file.path(data_dir, "panel.csv")) # Can also be .xlsx
metadata <- read_csv(file.path(data_dir, "metadata.csv"))

# Extract markers of interest
markers <- panel %>% 
  filter(Type != "none") %>% 
  pull(Antigen)

# Read FCS files
flowset <- compile_fcs(
  data_dir = data_dir,
  pattern = "\\.fsc" # Read all FCS files
)

# Convert flowset to tibble
df <- convert_flowset(
  flowset = flowset,
  metadata = metadata,
  filename_col = "Filename",
  sample_ids = "Filename", # By default the filename is used to get sample ids
  batch_ids = "batch",
  condition = "condition",
  down_sample = TRUE,
  sample_size = 4000000,
  seed = 101,
  panel = panel, # Can also be the filename. It is solely used to ensure the channel names match what you expect (i.e. what is in the panel_antigen column)
  panel_channel = "Channel",
  panel_antigen = "Antigen"
)

# Transform data - This function also de-randomizes the data
uncorrected <- transform_asinh(
  df = df,
  markers = markers,
  cofactor = 5,
  .keep = TRUE # Lets you keep all columns, in case they are useful to you
)
# Store result
saveRDS(uncorrected, file = file.path(data_dir, "uncorrected.RDS"))
```

<!-- #### Minimal input use-case 

This example illustrates the minimal, runnable use-case.
Feel free to skip.
It is recommended to as a minimum include the batches from the metadata, either as above or as a vector of the same length as the data.


```{r, eval=FALSE}
# Load packages
library(cyCombine)
library(tidyverse)

# Directory with FCS files
data_dir <- "data"

# Extract markers from panel
panel_file <- "data/panel.csv" # Can also be .xlsx
metadata <- read.csv("data/metadata.csv") # Can also be .xlsx

# Extract markers of interest
markers <- read.csv(panel_file) %>% 
  filter(Type != "none") %>% 
  pull(Antigen)

# Prepare a tibble from directory of FCS files
uncorrected <- prepare_data(
  data_dir = data_dir,
  down_sample = FALSE
)

# Extract batches

  

# Store result
saveRDS(uncorrected, file = "data/uncorrected.RDS")
``` 
-->
<br><br>

### Batch correct

#### The recommended method

```{r, eval=FALSE}
# Load packages
library(cyCombine)
library(tidyverse)

# Load data (if not already loaded)
# uncorrected <- readRDS("data/uncorrected.RDS")
# markers <- get_markers(uncorrected)

# Batch correct
corrected <- batch_correct(
  df = uncorrected,
  covar = "condition",
  markers = markers,
  norm_method = "scale", # "rank" is recommended when combining data with heavy batch effects
  som_type = "fsom" # "som" is recommended when combining distinct data with heavy batch effects
)

# Save result
saveRDS(corrected, file.path(data_dir, "corrected.RDS"))
```

<br>

#### Modular method

```{r, eval = FALSE}
# Load packages
library(cyCombine)
library(tidyverse)

# Load data (if not already loaded)
# uncorrected <- readRDS("data/uncorrected.RDS")
# markers <- get_markers(uncorrected)

# Create cell type labels using a SOM grid (if you want to use your own labels, they can be added manually and this step should not be run)
labels <- uncorrected %>% 
  normalize(markers = markers,
            norm_method = "rank", # "scale" is recommended in cases with light batch effects (e.g. when combining similar data)
            ties.method = "average") %>% # Can also be minimum
  create_som(markers = markers, # Consider using create_fsom with norm_method = "scale", when combining similar data
             seed = 101,
             xdim = 8,
             ydim = 8)

# Batch correct
corrected <- correct_data(
  df = uncorrected,
  label = labels # Add custom labels here, if desired
  covar = "condition",
  markers = markers,
  parametric = TRUE
  )

# Save result
saveRDS(corrected, file.path(data_dir, "corrected.RDS"))
```

