---
title: "Detection of batch effects"
author: "Christina Bligaard Pedersen"
date: "March 2, 2021"
output: 
  prettydoc::html_pretty:
    theme: hpstr
    highlight: vignette
vignette: >
  %\VignetteIndexEntry{Detection of batch effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette will introduce the use of the batch effect detection module of cyCombine. For demonstrated use cases, see our [online vignettes](https://biosurf.org/cyCombine).

<br>

### Pre-processing data

First, packages must be loaded.

```{r libraries, eval=FALSE}
library(cyCombine)
library(tidyverse)

```


<br>

Then, it is possible to load the cytometry data - we suggest using the cyCombine package, but the batch detection functions work on all data frames. 

We are now ready to load the CyTOF data. We have set up a panel file in csv format, so the correct information is extractable from there.

```{r loading data, eval=FALSE}
# Directory with raw .fcs files
data_dir <- "~/data"

# Reading the panel from a CSV
panel <- read_csv(paste0(data_dir, "/panel.csv"))

# Extracting the markers
markers <- panel %>%
  filter(Type != "none") %>%
  pull(Marker) %>%
  str_remove_all("[ _-]")

# Reading cytometry data and converting to tibble
expr <- prepare_data(data_dir = data_dir,
                     metadata = paste0(data_dir, "/metadata.csv"),
                     filename_col = "FCS_name",
                     batch_ids = "Batch",
                     condition = "Set",
                     derand = TRUE,
                     cofactor = 5,
                     markers = markers,
                     down_sample = FALSE)

```

<br>

If loading mass cytometry (CyTOF) data, we suggest using derandomization and cofactor = 5 (default). For flow cytometry data, a cofactor of 150 is usually suitable, and for spectral flow cytometry, we recommend using cofactor = 6000.


<br><br>

### Checking for batch effects
Now for inspection of the data. We have two different functions: `detect_batch_effect_express` and `detect_batch_effect`. See the [online vignette](https://biosurf.org/cyCombine_detect_batch_effects.html) for a thorough discussion regarding these. Here, we just demonstrate how they are run.

<br>

#### detect_batch_effect_express
Here, we demonstrate how to use the `detect_batch_effect_express` function.


```{r detect batch effects express, eval=FALSE}
detect_batch_effect_express(df = expr,
                            batch_col = 'batch',
                            downsample = 10000, 
                            out_dir = paste0(data_dir, '/batch_effect_check'))
```
<br>

This function prints some diagnostics to the screen and provides three plots to the directory specified by `out_dir`. An MDS plot, a set of density plots (per-marker, per-batch), and a plot showing the batch-batch Earth Mover's Distances per-marker. 

<br><br>

#### detect_batch_effect
Here, we demonstrate how to use the `detect_batch_effect` function.

```{r detect batch effects, eval=FALSE}
detect_batch_effect(df = expr,
                    batch_col = 'batch',
                    out_dir = paste0(data_dir, '/batch_effect_check'), 
                    seed = 434,
                    name = 'CyTOF data')

```
<br>

This function prints some diagnostics to the screen and also provides UMAP plots, which are saved in the directory specified by `out_dir`. These plots should assist in detecting potential batch effects.
